%!TEX root = ./report.tex
\section{Background}
\label{sec:background}
\subsection{Coq}
\todo{Revise this section}
Coq is an interactive theorem prover for managing formal proofs. It is a language that implements the program specification language Gallina. Due to Gallinas base in CiC (Calculus of Inductive Constructions)\todo{Insert ref to CiC}, Coq implements both higher-order logics and a richly-typed functional programming language.

\subsection{Imp}
\label{sec:background_imp}
As previously mentioned, Imp\footnote{Not to be confused with Irons IMP language or the Edinburgh IMP language.} is a simple imperative programming language studied in the electronic book ``Software Foundations'' and implemented in Coq. The language includes a selection of core features of widely used imperative programming languages such as C and Java, and certain properties about the language as a whole have been proven. This helps verification of programs written in Imp, as the precise definition of Imp can be used to formally prove that a certain program satisfies a particular specification. 

At its core, the language consists of arithmetic and boolean expressions, which can be assigned to variables. The value of these variables at a point of execution is represented by a state, which is a partially applied function $id \rightharpoonup value$\todo{Give a definition of id?}. The language has a command for assignment of variables, along with control-flow statements for conditionals, loops, and sequencing of commands. To be able to reason about these commands, assertions are used to make claims about a particular state of a program during execution. These assertions are defined as $state \rightharpoonup Prop$. Each of the command is specified by a Hoare rule. These are shown in Figure \ref{fig:hoare_rules_imp}.

\begin{figure}
\[
    \infrule[Assignment]{}{
      	\triple
      		{Q\subst a X} 
      		{\;\cassign {X} {a}\;}
	  		{Q}
    }
    \;\;\;\;\;\;\;\;
    \infrule[Skip]{}{
      	\triple
      		{P} 
      		{\; \mathbf{SKIP} \;}
	  		{P}
    }
\]

\[
    \infrule[Sequence]{
    	\triple
      		{P} 
      		{\;c1\;}
	  		{Q}
	  		\;\;
	  	\triple
      		{Q} 
      		{\;c2\;}
	  		{R}
    }{
      	\triple
      		{P} 
      		{\;c1;c2\;}
	  		{R}
    }
    \;\;\;\;\;\;\;\;
    \infrule[While]{
    	\triple
      		{P \land b} 
      		{\; c \;}
	  		{P}
    }{
      	\triple
      		{P} 
      		{\; \mathbf{WHILE}\;{b}\;\mathbf{DO}\;{c}\;\mathbf{END} \;}
	  		{P \land \neg b}
    }
\]
\[
    \infrule[If]{
    	\triple
      		{P \land b} 
      		{\; c1 \;}
	  		{Q}
	  	\;\;
	  	\triple
      		{P \land \neg b} 
      		{\; c2 \;}
	  		{Q}
    }{
      	\triple
      		{P} 
      		{\;\mathbf{IFB}\;{b}\;\mathbf{THEN}\;{c1}\;\mathbf{ELSE}\;{c2}\;\mathbf{FI} \;}
	  		{Q}
    }
\]
\caption{Hoare rules of the Imp language.}
\label{fig:hoare_rules_imp}
\end{figure}
It is important to note that the Imp language investigated in this project is the one defined prior and up to the Hoare Logic chapter in ``Software Foundations''\footnote{http://www.itu.dk/courses/SASP/F2013/Hoare.html}, so extensions to this such as the type checker is not included. 


 \input{separation_logic}
